The ChEMBL database as Linked Open Data

### 2012-07-26 article.tex
Fixed the caption and png location
Addressed the comments from Antony Williams

\begin{abstract}
\paragraph*{Background:}
Making data available as Linked Data using Resource Description Framework (RDF) promotes integration with other web resources.
RDF documents can natively link to related data, and others can link back using Universal Resource Identifiers (URIs).
RDF makes the data machine-readable, using extensible vocabularies for additional information, making it easier
to scale up inference and data analysis.
\paragraph*{Results:}
This paper describes the continued project to convert data from the ChEMBL database into RDF triples.
This updated version of ChEMBL-RDF now uses recently introduced ontologies, including CHEMINF and CiTO,
exposes more information from the database, and is now available as dereferencable, linked data.
To demonstrate these new features, we present novel use cases showing further integration with
other web resources, including Bio2RDF, Chem2Bio2RDF, and ChemSpider, and showing the use of standard
ontologies.
\paragraph*{Conclusions:}
The use cases show that by using open standards and ontologies it is possible to link the ChEMBL database
to other databases. Using those links data integration can understood by software and create a foundation
for integrative cheminformatics applications, such as the presented decision support.
\end{abstract}

@@

The current scientific data deluge, in which datasets grow faster in size and complexity than scientists can handle,
provides several new challenges for information systems. Scientists wish to 
discover new, unique, and significant patterns in datasets, which explaining biological 
phenomena not yet understood. This process requires integration of datasets that are both 
sparse, and growing independent of one another. The discovery of new patterns showing the 
causes and effects of various biological phenomena is beyond the scope of single datasets.
For example, systems biologists integrate micro-array differential expression datasets to 
biological pathways, using various other datasets to provide evidence for the links~\cite{Staab2007}. 
Another prominent example is drug discovery where a new unique chemical entity is discovered 
based on descriptions of its biological properties. This process requires the effective 
linkage of many scientific datasets~\cite{Samwald2011,Williams2012}.

@@

Using the right to download, modify, and redistribute the ChEMBL data, two independent teams
have previously mapped  the ChEMBL dataset to the Resource Description Framework (RDF),
Chem2Bio2RDF~\cite{Chen2010} and ChEMBL-RDF~\cite{Willighagen2011}. RDF is a framework where
knowledge is represented by small, so-called triples reflecting very basic facts. Things
in RDF are represented by Universal Resource Identifiers (URI) and when the more specialized
Universal Resource Locator (URL) is used, it creates \emph{linked data}~\cite{Samwald2011}. RDF is
developed as an open standard by the World Wide Web Consortium, and provides a knowledge
representation framework that applies to many, if not any, domain, making it a perfect
approach for data integration in the life sciences.

This paper presents an update on the ChEMBL-RDF dataset, and details of the latest structures
and ontologies used to map ChEMBL 13 to RDF, along with new use cases, showing 
links to other datasets using their RDF URIs to further support research in the life sciences.

@@

The ChEMBL version used in this paper is ChEMBL 13, which was released on 29 February 2012.
The SQL data dump provided by ChEMBL was inserted into a local MySQL database. A set of SQL queries 
were then executed to generate RDF files using a set of PHP scripts. These scripts are Open Source 
and are available from the source code hosting service, GitHub~\cite{ChEMBLRDFGitHub}. This process 
has been outlined in a best practices note by the W3C Health Care and Life Sciences Interest
Group~\cite{Marshall2012}. This process has not changed much.

The previous ChEMBL-RDF conversion used, however, custom predicates and classes to represent RDF triples, defining
an ad-hoc ontology~\cite{Willighagen2011}. The current version, instead, uses community-proposed
ontologies, making the RDF more interoperable. The new triple dataset uses standard ontologies
including the Bibliography Ontology~\cite{Giasson2011} and the Citation Typing Ontology~\cite{Hastings2011} for literature
references and citations, and domain ontologies like the Protein Ontology\cite{Sidhu2006} and the Chemical Information
Ontology\cite{Hastings2011}. Throughout this paper, various prefixes are used to denote different ontology 
namespaces to simplify the RDF examples. These prefixes are outlined in Table~\ref{namespaces}.

@@

To expose that ChEMBL-RDF data two approaches have been adopted. First, a SPARQL end point
hosted at Uppsala University, using the Open Source Virtuoso software. Use is free, but the
querying is capped, based on the estimated computational effort. Second, resources have
been made dereferencable via the \url{http://linkedchemistry.info} resource. The Kasabi
platform~\cite{kasabi} was used for a period of time, until that project was discontinued.
The Uppsala SPARQL end point still uses resource IRIs based on this, and follow a similar
IRI pattern where linkedchemistry.info/chembl is replaced by data.kasabi.com/dataset/chembl-rdf.

@@

We here present the updated ChEMBL-RDF, including a description of the mapping process 
and the identification of links to other datasets.

@@

The name and synonyms for each drug are
provided as labels. When InChI and InChIKeys are available for a drug, then these are
provided via the CHEMINF formalism (here abbreviated):

@@

Similarly, properties of small molecules are available from ChEMBL, and as of ChEMBL-RDF 13
these too are exposed in triple format. Like the InChI and InChIKey, these are provided using the
CHEMINF ontology approach. Here is, for example, the ALogP value:

@@

The literature from which all this data was extracted, is replicated from the database,
but only a subset of information about that literature is provided. In all cases,
it uses the PubMed Identifiers (PMIDs) and Digital Object Identifiers (DOIs) to link
to authoritative databases to provide full detail, as outlined later in this paper.
The provided properties are using the BIBO ontology:

@@

Each release of the ChEMBL database is accompanied by set of release counts, which help 
provide a concise content overview of the release in question. The ChEMBL 13 release counts
are as follows:
\begin{itemize}
  \item 8,845 targets
  \item 1,143,682 distinct molecules
  \item 6,933,068 activities
  \item 617,681 assays
  \item 44,682 documents
\end{itemize}

@@

The following SPARQL query returns the 44,681 ChEMBL documents. This is 1 less than appears 
in ChEMBL 13 release notes, as it does not include doc\_id -1, which refers to the 
unpublished dataset.

@@

Recreation of the ChEMBL release statistics, by querying the ChEMBL-RDF with SPARQL 
provides end users with high degree of confidence that the ChEMBL-RDF contains same 
core content as source relational database.

@@

ChemSpider~\cite{Pence2010} is a freely accessible, online database provided by Royal Society of Chemistry (RSC).
It contains over 26 million unique
chemical compounds aggregated from over 400 data sources as well as chemical data extracted
from RSC scientific articles and databases. Since its inception, efforts have been made to
utilize it as both a deposition platform for the community to contribute novel data, as
well as a platform for annotation and curation for existing data. Studies have shown that there are data
quality issues in many of the public compound databases~\cite{Williams2011}. ChemSpider has become a
valuable resource for curated data, especially chemical-compound name mappings. ChemSpider
is presently providing the chemical structure, substructure and similarity searching services 
underpinning the Open PHACTS semantic web project~\cite{Williams2012}. Specific chemical data 
sources containing data mappings between ChemSpider identifiers (CSIDs) and the original data 
source identifiers have been provided to the triple store, together with chemical identifiers 
including validated chemical names (systematic, generic and trivial), SMILES, and InChIs. 
The data mappings between the CSIDs and ChEMBL IDs are released to the 
community under the Creative Common Attribution-Share Alike license (CC-BY-SA 3.0). Attribution 
should be made to the original ChEMBL database, Open PHACTS, and ChemSpider.

@@

Notice here the use of owl:equivalentClass to match the formalism in CHEMINF that defines molecules
are classes, rather than instances.

@@

The links services enables the ChEMBL application to derive both forward links, originating in
ChEMBL, e.g. \url{http://localhost:8080/chembl/linkstonamespace/-targetns/originalns:identifier},
and backward links, originating in other databases, such as LODD, Bio2RDF,
\footnote{\url{http://bio2rdf.org/linksns/targetns/originalns:identifier}}, and Chem2Bio2RDF.
These services are vital to efficiently navigate the Linked Data web, as it is both
impractical and inefficient to require users to crawl the entire web before they can discover 
relevant resources. These services are currently only supplied as web services from ChEMBL and
Bio2RDF, but it is hoped that similar services will be provided by other scientific Linked Data
providers in the future. Datasets that are available in SPARQL endpoints can be queried for
links efficiently using simple queries as demonstrated in the ChEMBL web application.

@@

A second resource in the linked open data network in which ChEMBL-RDF has been integrated is Chem2Bio2RDF.
Chem2Bio2RDF created a single RDF repository covering over 20 public data resources pertaining to drugs, chemical compounds, carcinogens,
protein targets, genes, diseases, side effects, pathways and their relations~\cite{Chen2010}. The entities and their relations were
further annotated by Chem2Bio2OWL ontology, making it a rich semantic resource for integrative search~\cite{Chen2012a} and
mining~\cite{Chen2012b}.  ChEMBL-RDF set (version 13) has been uploaded into Chem2Bio2RDF triple store, enabling to perform
queries linking ChEMBL with other entities in Chem2Bio2RDF. Since both ChEMBL RDF and Chem2Bio2RDF use InChI keys to present chemical
compounds and adopt Bio2RDF protein identifiers to present targets, queries can be easily constructed to link each other. For instance,
to investigate the relations between drug side effects and their targets (usually off-targets), a query was created to link the
targets in ChEMBL to side effects (e.g., heart disease) in Chem2Bio2RDF via side effect related drugs and their bioassay activities.
In this case, 36 drugs causing heart diseases were linked to 87 unique protein targets (IC$_50$ $<$ 10um). The top two most common targets
of these drugs are HERG and Serotonin transporter. HERG is a well-known off-target accounting for cardiovascular problems, while
the activity of Serotonin transporter may also be associated with significant cardiovascular adverse effects~\cite{Levy2006}. 
The targets can be also linked to side effects via protein protein interactions by searching across ChEMBL, HPRD and Sider in
Chem2Bio2RDF. SPARQLs are available at \url{http://chem2bio2rdf.wikispaces.com/Integrate+with+ChEMBL} and can be executed in the
SPARQL endpoint~\cite{Chen2010}.

\begin{figure*}[!ht]
		\begin{center}
		\includegraphics[width=15cm]{cited.png}
		\newline
		\caption[wee]{Screenshot from Bioclipse Decision Support with results from a ChemSpider + ChEMBL-RDF search. The top left canvas contains the query structure, in this case the drug Carbamezapine, the top right canvas shows the near neighbors in ChemSpider (via a similarStructure search) that contain ChEMBL-RDF data, the lower right shows the chemical structure for the selected compound in the top right canvas, and the lower left canvas shows the found interactions for this compound using ChEMBL-RDF.}
	\label{fig:bioclipse-ds}
		\end{center}
\end{figure*}

@@

Using the SPARQL end point, it is easy to integrate data from ChEMBL into other web resources, such as in CitedIn, (http://www.citedin.org).
The impact of a scientific publication is traditionally measured by frequency of citation in subsequent publications, this narrowly defined calculation increasingly fails to capture the true reach of a paper. CitedIn was developed to make scientific impact more visible and more accurate by providing insight into where published work is being cited online in a variety of public resources. ChEMBL contains many literature references. The SPARQL endpoint is queried by CitedIn and the results are presented to its users. Figure shows the result for the Nature paper by Gamo presenting the malari structures, which is highly cited in ChEMBL~\cite{Gamo2010}. Counting citations in ChEMBL is done by the following SPARQL query.

@@

NOTE: May want to exclude this Use Case as query has not successful run yet, keeps causing virtuoso 
to timeout or run out of memory. Currently playing with local server setup to get query to run.

Designing a molecule, which is selective to one target over another will often be considered a 
successful outcome in a drug design process. If the target in question is a protein, it is easy
to understand that as sequence identity amongst residues, which contribute to potential small 
molecule binding sites remains high so does the issue of selectivity. That is, the molecule being 
designed may also bind to the equivalent binding site in the  closely related protein 
target and may lead to undesirable consequences. The ChEMBL data model links molecules to 
targets using different activity types recorded in the literature. Using activity types, 
which act as a measure of binding affinity, such as IC$_{50}$ or K$_{i}$ and applying activity value cuts
offs it is possible to identify molecules which have higher binding affinity to certain targets 
compared to others. Taking this one step further it is possible to identify a set molecules, 
which have a high affinity to protein A (e.g. IC$_{50}$ value $<$ 50~nM) and low affinity to protein 
B (e.g. IC$_{50}$ value $>$ 200~nM).

@@

This papers shows the evolution of the ChEMBL-RDF resource in standardizing access to the
ChEMBL data in a machine-readable manner, by using semantic web technologies. This standardization
applies to several aspects of data exchange, including the data serialization format and
access options, the data structure, and the ontologies used to explain what the data is about.
We touched upon choices made to implement these approaches, such as a selection of
RDF technologies, and chosen ontologies.

For example, for linking to other databases we selected a few predicates to link to
other databases: rdfs:seeAlso, skos:exactMatch, owl:equivalentClass, owl:sameAs. Each
has an increasing level of implications, ranging from "you may also be interested in" to
"both objects are identical and all properties apply to both objects". Additionally,
the CiTO ontology has been used to link data to the literature it was extracted from,
which provides link that lies in between rdfs:seeAlso and skos:exactMatch.

@@

Finally, we stress that nothing in this paper cannot be done with other approaches, such as
more common relational databases. In fact, that is exactly what the ChEMBL website does.
Similarly, there is nothing principle component analysis does that you cannot do with drawings
on paper too; it is just silly. We have here presented a linked data approach to the ChEMBL data
showing standardized and simplified integration to other resources, and that is what
cheminformatics is about.

\section*{Authors contributions}
EW initiated the project, created the initial RDF version of the ChEMBL data, and encouraged the use cases.
AW extended CitedIn to support citation info in ChEMBL-RDF and supported setting up the http://linkedchemistry.info/ host.
OS, AW, and VT developed the nearest neighbor application of Bioclipse and ChemSpider.
JH supported the project with CHEMINF representations.
PA integrated ChEMBL-RDF into Bio2RDF.
JO, AG, and MD critically validated and summarized the ChEMBL data statistics.
All authors contributed to the continued development of ChEMBL-RDF, as well as to the writing of the paper,
and approved the final version.

>>>>

\begin{abstract}
\paragraph*{Background:}
Making data available as Linked Data using Resource Description Framework (RDF) promotes integration with other web resources.
RDF documents can natively link to related data, and others can link back using Universal Resource Identifiers (URIs).
RDF makes the data machine-readable, using extensible vocabularies for additional information, making it easier
to scale up inference and data analysis.
\paragraph*{Results:}
This paper describes the continued project to convert data from the ChEMBL database into RDF triples.
This updated version of ChEMBL-RDF now uses recently introduced ontologies, including CHEMINF and CiTO,
exposes more information from the database, and is now available as dereferencable, linked data.
To demonstrate these new features, we present novel use cases showing further integration with
other web resources, including Bio2RDF, Chem2Bio2RDF, and ChemSpider, and showing the use of standard
ontologies.
\paragraph*{Conclusions:}
The use cases show that by using open standards and ontologies it is possible to link the ChEMBL database
to other databases. Using those links data integration can understood by software and create a foundation
for integrated cheminformatics applications, such as the presented decision support.
\end{abstract}

@@

The current scientific data deluge, in which datasets grow faster in size and complexity than scientists can handle,
provides several new challenges for information systems. Scientists wish to 
discover new, unique, and significant patterns in datasets that explain biological 
phenomena not yet understood. This process requires integration of datasets that are both 
sparse, and growing independently of one another. The discovery of new patterns showing the 
causes and effects of various biological phenomena is beyond the scope of single datasets.
For example, systems biologists integrate micro-array differential expression datasets to 
biological pathways, using various other datasets to provide evidence for the links~\cite{Staab2007}. 
Another prominent example is drug discovery where a new unique chemical entity is discovered 
based on descriptions of its biological properties. This process requires the effective 
linkage of many scientific datasets~\cite{Samwald2011,Williams2012}.

@@

Using the right to download, modify, and redistribute the ChEMBL data, two independent teams
have previously mapped  the ChEMBL dataset to the Resource Description Framework (RDF),
Chem2Bio2RDF~\cite{Chen2010} and ChEMBL-RDF~\cite{Willighagen2011}. RDF is a framework where
knowledge is represented by small so-called triples reflecting very basic facts. Any fact can
be given on anything, and RDF therefore talks about \textit{things}~\cite{Miller:04:RP}. Things in RDF
are represented by Universal Resource Identifiers (URI) and when the more specialized
Universal Resource Locator (URL) is used, it creates \emph{linked data}~\cite{Samwald2011}. RDF is
developed as an open standard by the World Wide Web Consortium, and provides a knowledge
representation framework that applies to many, if not any, domain, making it a perfect
approach for data integration in the life sciences.

This paper presents an update on the ChEMBL-RDF dataset, and details of the latest structures
and ontologies used to map ChEMBL to RDF, along with new use cases, showing 
links to other datasets using their RDF URIs to further support research in the life sciences.

@@

The ChEMBL version used in this paper is ChEMBL 13, which was released on 29 February 2012.
The SQL data dump provided by ChEMBL was inserted into a local MySQL database. A set of SQL queries 
were then executed to generate RDF files using a set of PHP scripts. These scripts are Open Source 
and are available from the source code hosting service, GitHub~\cite{ChEMBLRDFGitHub}. This process 
has been outlined in a best practices note by the W3C Health Care and Life Sciences Interest
Group~\cite{Marshall2012}. This process has not changed much since its original description.

However, The previous ChEMBL-RDF conversion used custom predicates and classes to represent RDF triples, defining
an ad-hoc ontology~\cite{Willighagen2011}. The current version, instead, uses community-proposed
ontologies, making the RDF more interoperable. The new triple dataset uses standard ontologies
including the Bibliography Ontology~\cite{Giasson2011} and the Citation Typing Ontology~\cite{Hastings2011} for literature
references and citations, and domain ontologies like the Protein Ontology\cite{Sidhu2006} and the Chemical Information
Ontology\cite{Hastings2011}. Throughout this paper various prefixes are used to denote different ontology 
namespaces to simplify the RDF examples. These prefixes are outlined in Table~\ref{namespaces}.

@@

To expose the ChEMBL-RDF data two approaches have been adopted. First, a SPARQL end point
hosted at Uppsala University, using the Open Source Virtuoso software. Use is free, but the
querying is capped, based on the estimated computational effort. Second, resources have
been made dereferencable via the \url{http://linkedchemistry.info} resource. The Kasabi
platform~\cite{kasabi} was used for a period of time, until that project was discontinued.
The Uppsala SPARQL end point still uses resource IRIs based on this, and follow a similar
IRI pattern where linkedchemistry.info/chembl is replaced by data.kasabi.com/dataset/chembl-rdf.

@@

We present here the updated ChEMBL-RDF, including a description of the mapping process 
and the identification of links to other datasets.

@@

The name and synonyms for each drug are
provided as labels. When InChI and InChIKeys are available for a drug, then these are
provided via the CHEMINF formalism (abbreviated here):

@@

Similarly, properties of small molecules are available from ChEMBL, and as of ChEMBL-RDF 13
these too are exposed in triple format. Like the InChI and InChIKey, these are provided using the
CHEMINF ontology approach. Here is, for example, the ALogP value (as distributed with ChEMBL and
calculated with ACD/Labs software):

@@

The literature from which all this data was extracted, is replicated from the database,
but only a subset of information about that literature is provided. In all cases
it uses the PubMed Identifiers (PMIDs) and Digital Object Identifiers (DOIs) to link
to authoritative databases to provide full detail, as outlined later in this paper.
The provided properties are using the BIBO ontology:

@@

Each release of the ChEMBL database is accompanied by a set of release counts, which help 
provide a concise content overview of the release in question. The ChEMBL 13 release counts
are as follows:
\begin{itemize}
  \item 8,845 targets
  \item 1,143,682 distinct molecules
  \item 6,933,068 activities
  \item 617,681 assays
  \item 44,682 documents
\end{itemize}

@@

The following SPARQL query returns the 44,681 ChEMBL documents. This is 1 less than appears 
in ChEMBL 13 release notes~\cite{ChEMBL13ReleaseNotes}, as it does not include doc\_id -1, which refers to the 
unpublished dataset.

@@

Recreation of the ChEMBL release statistics, by querying the ChEMBL-RDF with SPARQL 
provides end users with a high degree of confidence that the ChEMBL-RDF contains the same 
core content as the source relational database.

@@

ChemSpider~\cite{Pence2010} is a freely accessible, online database provided by Royal Society of Chemistry (RSC).
It contains over 28 million unique
chemical compounds aggregated from over 400 data sources as well as chemical data extracted
from RSC scientific articles and databases. Since its inception, efforts have been made to
utilize it as both a deposition platform for the community to contribute novel data, as
well as a platform for annotation and curation for existing data. Studies have shown that there are data
quality issues in many of the public compound databases~\cite{Williams2011}. ChemSpider has become a
valuable resource for curated data, especially chemical-compound name mappings. ChemSpider
is presently providing the chemical structure, substructure and similarity searching services 
underpinning the Open PHACTS semantic web project~\cite{Williams2012}. Specific chemical data 
sources containing data mappings between ChemSpider identifiers (CSIDs) and the original data 
source identifiers have been provided to the triple store, together with chemical identifiers 
including validated chemical names (systematic, generic and trivial), SMILES, and InChIs. 
The data mappings between the CSIDs and ChEMBL IDs are released to the 
community under the Creative Common Attribution-Share Alike license (CC-BY-SA 3.0). Attribution 
should be made to the original ChEMBL database, Open PHACTS, and ChemSpider.

@@

Notice here the use of owl:equivalentClass to match the formalism in CHEMINF that defines molecules
as classes, rather than instances.

@@

These links service enables the ChEMBL application to derive both forward links, originating in
ChEMBL, e.g. \url{http://localhost:8080/chembl/linkstonamespace/-targetns/originalns:identifier},
and backward links, originating in other databases, such as LODD, Bio2RDF
(\url{http://bio2rdf.org/linksns/targetns/originalns:identifier}), and Chem2Bio2RDF.
These services are vital to efficiently navigate the Linked Data web, as it is both
impractical and inefficient to require users to crawl the entire web before they can discover 
relevant resources. These services are currently only supplied as web services from ChEMBL and
Bio2RDF, but it is hoped that similar services will be provided by other scientific Linked Data
providers in the future. Datasets that are available in SPARQL endpoints can be queried for
links efficiently using simple queries as demonstrated in the ChEMBL web application.

@@

A second resource in the linked open data network in which ChEMBL-RDF has been integrated is Chem2Bio2RDF.
Chem2Bio2RDF created a single RDF repository covering over 20 public data resources pertaining to drugs, chemical compounds, carcinogens,
protein targets, genes, diseases, side effects, pathways and their relations~\cite{Chen2010}. The entities and their relations were
further annotated by Chem2Bio2OWL ontology, making it a rich semantic resource for integrative search~\cite{Chen2012a} and
mining~\cite{Chen2012b}. The ChEMBL-RDF set has been uploaded into the Chem2Bio2RDF triple store, enabling
queries linking ChEMBL with other entities in Chem2Bio2RDF. Since both ChEMBL RDF and Chem2Bio2RDF use InChI keys to present chemical
compounds and adopt Bio2RDF protein identifiers to present targets, queries can be easily constructed to link each other. For instance,
to investigate the relations between drug side effects and their targets (usually off-targets), a query was created to link the
targets in ChEMBL to side effects (e.g., heart disease) in Chem2Bio2RDF via side effect related drugs and their bioassay activities.
In this case, 36 drugs causing heart diseases were linked to 87 unique protein targets (IC$_50$ $<$ 10um). The top two most common targets
of these drugs are HERG and Serotonin transporter. HERG is a well-known off-target accounting for cardiovascular problems, while
the activity of Serotonin transporter may also be associated with significant cardiovascular adverse effects~\cite{Levy2006}. 
The targets can be also linked to side effects via protein-protein interactions by searching across ChEMBL, HPRD and Sider in
Chem2Bio2RDF. SPARQLs are available at \url{http://chem2bio2rdf.wikispaces.com/Integrate+with+ChEMBL} and can be executed in the
SPARQL endpoint~\cite{Chen2010}.

\begin{figure*}[!ht]
		\begin{center}
		\includegraphics[width=15cm]{citedin.png}
		\newline
		\caption[wee]{Screenshot of the CitedIn web service showing a paper cited 94671 times in the ChEMBL database.}
	\label{fig:citedin}
		\end{center}
\end{figure*}

@@

Using the SPARQL end point, it is easy to integrate data from ChEMBL into other web resources, such as in CitedIn, (http://www.citedin.org).
The impact of a scientific publication is traditionally measured by frequency of citation in subsequent publications, this narrowly defined calculation increasingly fails to capture the true reach of a paper. CitedIn was developed to make scientific impact more visible and more accurate by providing insight into where published work is being cited online in a variety of public resources. ChEMBL contains many literature references. The SPARQL endpoint is queried by CitedIn and the results are presented to its users. Figure~\ref{fig:citedin} shows the result for the Nature paper by Gamo presenting the malaria structures, which is highly cited in ChEMBL~\cite{Gamo2010}. Counting citations in ChEMBL is done by the following SPARQL query.

@@

Designing a molecule, which is selective to one target over another will often be considered a 
successful outcome in a drug design process. If the target in question is a protein, it is easy
to understand that as sequence identity amongst residues, which contribute to potential small 
molecule binding sites remains high, so does the issue of selectivity. That is, the molecule being 
designed may also bind to the equivalent binding site in the  closely related protein 
target and may lead to undesirable consequences. The ChEMBL data model links molecules to 
targets using different activity types recorded in the literature. Using activity types, 
which act as a measure of binding affinity, such as $IC_{50}$ or $K_{i}$ and applying activity value cuts
offs it is possible to identify molecules which have higher binding affinity to certain targets 
compared to others. Taking this one step further it is possible to identify a set molecules
which have a high affinity to protein A (e.g. $IC_{50}$ value $<$ 50~nM) and low affinity to protein 
B (e.g. $IC_{50}$ value $>$ 200~nM).

@@

This paper shows the evolution of the ChEMBL-RDF resource in standardizing access to the
ChEMBL data in a machine-readable manner, by using semantic web technologies. This standardization
applies to several aspects of data exchange, including the data serialization format and
access options, the data structure, and the ontologies used to explain what the data is about.
We touched upon choices made to implement these approaches, such as a selection of
RDF technologies, and chosen ontologies.

For example, for linking to other databases we selected a few predicates to link to
other databases: rdfs:seeAlso, skos:exactMatch, owl:equivalentClass, owl:sameAs. Each
has an increasing level of implications, ranging from "you may also be interested in" to
"both objects are identical and all properties apply to both objects". Additionally,
the CiTO ontology has been used to link data to the literature it was extracted from,
which provides links that lie in between rdfs:seeAlso and skos:exactMatch.

@@

Finally, we stress that nothing in this paper cannot be done with other approaches, such as
more common relational databases. In fact, that is exactly what the ChEMBL website does.
Similarly, there is nothing principle component analysis does that you cannot do with drawings
on paper too; it is just inappropriate to do so when computer-based approaches are so much
more efficient. We have presented here a linked data approach to the ChEMBL data
showing standardized and simplified integration to other resources, and that is what
cheminformatics is about.

\section*{Authors contributions}
EW initiated the project, created the initial RDF version of the ChEMBL data, and encouraged the use cases.
AW extended CitedIn to support citation info in ChEMBL-RDF and supported setting up the http://linkedchemistry.info/ host.
OS, AJW, and VT developed the nearest neighbor application of Bioclipse and ChemSpider.
JH supported the project with CHEMINF representations.
PA integrated ChEMBL-RDF into Bio2RDF.
JO, AG, and MD critically validated and summarized the ChEMBL data statistics.
All authors contributed to the continued development of ChEMBL-RDF, as well as to the writing of the paper,
and approved the final version.
